---
title: "scRNASeq analysis of cycling cardiomyocytes"
author: "Jovinge Lab"
date: "5/16/2018"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "/mnt/data/scratch/christy/AGG/outs")

font_family <- "Times New Roman"
library(extrafont)
loadfonts()
```

## Introduction

This document was prepared in support of our paper:

<center>
<b>Isolation of cardiomyocytes undergoing mitosis with complete cytokinesis.</b>

Hsiao-yun Y. Milliron Ph.D1†, Matthew J. Weiland M.S.1†, Eric J. Kort M.D. M.S.1,2†, and Stefan Jovinge MD Ph.D1,3,4*

†Equal contribution

</center>

This document documents how the gene expression data was analyzed and the figures produced or 
this manuscript. The source file (`data_analysis.rmd`) for this document and 
supporting data files are available from 
[the project github repository](https://github.com/vanandelinstitute/cycling). 
Compiling (kniting) the source document in RStudio (or from the command line) 
will regenerate key panes off figures 5 and 6 from the source data as they appear in the manuscript. Note that due to the large data files, cloning the repository 
requires that you have [large file support for git](https://git-lfs.github.com) 
installed.  Otherwise, you will only receive the soft links to the data files, not the data files themselves.

## Pre-requisites

This analysis requires a number of packages to complete, most notably AUCell, scde, and 
cellRangerRkit (which is no longer supported by 10XGenomics, but is still available on 
the internet). These dependencies may be installed as follows. First some system libraries must be installed from the command line.

```{bash, eval=FALSE}

# Note that scde relies on Cairo, which in turn relies on the cairo and 
# libxt dev files.
# on debian they may be installed as:
sudo apt-get install libcairo2-dev
sudo apt-get install libxt-dev

```

Then the required R packages can be installed within an R session.

```{r, eval=FALSE}
devtools::install_github("hb-gitified/cellrangerRkit")
devtools::install_github("hms-dbmi/pagoda2")

setRepositories(ind=c(1,2,3,4))
install.packages("scde")
install.packages("AUCell")
install.packages("doRNG")
install.packages("extrafont")

```

The required packages can then be loaded. If the following fails with an 
error, additional libraries may be missing on your system but should be 
readily installed with `install.packages(...)`.


```{r, eval=TRUE}

loadAll <- function() {
  library(reshape2)
  library(org.Hs.eg.db)
  library(cellrangerRkit)
  library(scde)
  library(DOSE)
  library(clusterProfiler)
  library(AUCell)
  library(pagoda2)
  library(ggplot2)
  library(ggpubr)
  library(grid)
  library(gridExtra)
}

suppressMessages(loadAll())
```

## Background
Pharmacologic and interventional advances over the past two decades have
produced a substantial reduction in mortality related to coronary artery disease
(CAD).  However, while mortality is on the decline, prevalance of the sequelae
of CAD--most notably dilated cardiomyopathy and heart failure--are on the rise.
Restoration of function among patient experiencing ischemia or infarct related
to CAD requires replenishment of cardiomyocytes within the failing myocardium.
Our past work has demonstrated that the heart has only limited regenerative
capacity in this regard.  Therefore, it is of utmost importance to be able to
identify those cardiomyocytes with proflierative potential, either for the
purposes of local delivery of these cells or to identify new pharmacologic
strategies for stimulating cardiomyocyte profiferation.

Conventional markers of cell cycle progression based on DNA content are not
useful for identifying proliferating cardiomoycytes.  This is due to the fact
that while cardiomyocytes rarely complete cell division, they commonly enter the
cell cycle but arrest prior to cytokinesis.

Here we combine molecular beacon technology with single cell RNASeq profiling to
identify and chracterize a population of cardimoyocytes enriched for cells with
proliferative potential.


## Differential gene expression analysis
Here we use the `pagoda2` package to normalize the gene counts and then identify
genes that are (1) differentially expressed between each group of cells vs. the
other groups and (2) also on average more highly expressed in each group vs.
both the other groups.  We will restrict our analysis to cells that are TMEM88
or TNNT2 positive (cardiomyocytes and/or cardiomyocyte progenitors).

The code below shows how the raw counts were extracted from the 10X Chromium run 
and how they were filtered. The raw output from the 10X Chromium pipeline is not 
provided here due to file size, but this repository contains the filtered raw 
counts for use in both the pagoda and AUCell analyses below.

```{r, echo=TRUE, eval=FALSE}

root <- "./final_data/outs/"

countMatrix <- read.10x.matrices(
  paste0(root, "filtered_gene_bc_matrices_mex/GRCh38/")
)
cd <- clean.counts(as.matrix(countMatrix), 
                   min.lib.size=1000, 
                   min.reads = 1, 
                   min.detected = 1)
cd <- apply(cd, 2, function(x) { storage.mode(x) <- 'integer'; x})
colnames(cd) <- gsub("^.*_", "", colnames(cd))

# restrict to cardiomyocytes or cardiomyocyte precursors
ix <-  which(rownames(cd) == "TMEM88")[1]
ix2 <-  which(rownames(cd) == "TNNT2")[1]
cd <- cd[, which(cd[ix,] > 0 | cd[ix2,] > 0)]

# these raw counts are used later by AUCell
saveRDS(cd, "final_data/filtered_count_matrix.rds")

#normalize and filter using pagoda defaults
counts <- gene.vs.molecule.cell.filter(cd)
#counts <- counts[ rowSums(as.matrix(counts)) >= 10,]
rownames(counts) <- make.unique(rownames(counts))
saveRDS(counts, file="final_data/counts.rds")
pag <- Pagoda2$new(counts, log.scale = TRUE, n.cores = 8)
pag$adjustVariance(plot = FALSE, gam.k = 10)
pag <- saveRDS(pag, file="final_data/pagoda_object.rds")
```

We can then load the cell annotations (cell cycle and CDC level as 
quantified by molecular beacons) from the provided metadata file 
and identify differentially expressed genes between the cell populations.

The resulting pagoda data object is saved to file for future reference.

```{r, eval=FALSE, results='hide'}

# identify differentially expressed genes
pd <- readRDS("final_data/cycling_metadata.rds")
sg <- pd$type_cdc20
sg <- sg[which(pd$barcode %in% rownames(pag$counts))]
names(sg) <- rownames(pag$counts)
sg <- factor(sg)
sg2 <- pd$type
sg2 <- sg2[which(pd$barcode %in% rownames(pag$counts))]
sg2 <- factor(sg2)

pag$misc$groups <- sg
pag$getDifferentialGenes(groups = pag$misc$groups, upregulated.only = FALSE, z.threshold=0)
saveRDS(pag, file="final_data/pagoda_object.rds")

```

A volcano plot of this analysis was generated as follows, 
which is presented in the manuscript as supplemental figure 4a. 

```{r}
pag <- readRDS(file="final_data/pagoda_object.rds")
oe <- pag$diffgenes$counts$customClustering$G2M_pos
pv <- 2*pnorm(-abs(oe$Z))
pv.adj <- p.adjust(pv, "BH")

color <- rep("dark gray", nrow(oe))
color[oe$M < -0.5849625 & pv.adj < 0.001] <- "#00B2FF"
color[oe$M > 0.5849625 & pv.adj < 0.001] <- "red"
dat <- data.frame(logFC = oe$M, logP = -log10(pv.adj), color=color)

ggplot(dat, aes(x = logFC, y = logP, color = color)) +
  scale_colour_manual(values = c("#00B2FF", "dark gray", "red")) +
  geom_point(size = 1, alpha = 1, na.rm = T) + 
  scale_y_continuous(trans = "log1p") + 
  ylab(expression(-log[10]("FDR"))) +
  xlab(expression(log[2]("Fold Change"))) +
  geom_hline(yintercept = 3, colour="#990000", linetype="dashed") + 
  geom_vline(xintercept = 0.5849625, colour="#990000", linetype="dashed") + 
  geom_vline(xintercept = -0.5849625, colour="#990000", linetype="dashed") +
  xlim(-2, 2) +
  theme_bw() +
  theme(legend.position = 'none') 

ix <- which(oe$M > 0.5849625 & pv.adj < 0.001)
ix.d <- which(oe$M <= -0.5849625 & pv.adj <= 0.001)
```

The most significant 25 genes are listed in the table below.

```{r, results = "asis"}
knitr::kable(oe[1:25,], caption = "Genes up-regulated in CDC20+ cells")
```


## Functional Analysis

### Cell cycle gene enrichment

We can examine the functional the differentially expressed genes identified 
above by calculating enrichment in GO terms via the hypergeometric test. 
The most enriched terms are shown in the figure 5b & c, generated below.

```{r, echo = TRUE}


# note that using SYMBOLS causes enrichGO to fail (as it tries to convert 
# the SYMBOLS to SYMBOLS which results in fewer columns of gene ids than 
# the function expects).  So we convert to entrez gene id.

symToEg <- function(x) {
  ix <- which(x %in% ls(org.Hs.egSYMBOL2EG))
  eg <- unlist(mget(x[ix], org.Hs.egSYMBOL2EG))
  return(unique(eg))
}

bg <- colnames(pag$counts)
oe <- pag$diffgenes$counts$customClustering$G2M_pos
oe.sig.up <- rownames(oe[ix,])
oe.sig.down <- rownames(oe[ix.d,])


ego.u <- enrichGO(gene = symToEg(oe.sig.up), 
                OrgDb = org.Hs.eg.db, 
                ont = "BP", 
                pAdjustMethod = "BH", 
                qvalueCutoff = 0.05, 
                readable = TRUE)

pdf(file="fig5b.pdf", width = 2500/300, height=2500/300, family = font_family)
g <- dotplot(ego.u, showCategory=30)  
print(g)
dev.off()

ego.d <- enrichGO(gene = symToEg(oe.sig.down), 
                OrgDb = org.Hs.eg.db, 
                ont = "BP", 
                pAdjustMethod = "BH", 
                qvalueCutoff = 0.05, 
                readable = TRUE)
                
pdf(file="fig5c.pdf", width = 2800/300, height=1500/300, family = font_family)
g <- dotplot(ego.d, showCategory=30) 
print(g)
dev.off()

```

We can further analyze the functional difference in these cells by performing 
gene set enrichment analysis (via the AUCell algorithm) using established cell 
cycle signature. To begin, we define a plotting function for AUC scores to color 
by cell group and facet by signature.

```{r}
plot_auc <- function(data, title = " ", comparisons) {
  ttest2 <- function(a, b) {
    p <- wilcox.test(a,b)$p.value
    p <- ifelse(p < 0.0001, "<0.0001", round(p,4))
    p <- ifelse(p > 0.05, "ns", p)
    list(p.value=p)
  }
  
  print(paste0("Kruskal Wallis Test p-value: ", 
               kruskal.test(value ~ class, data=data)$p.value))
  
  p <- ggplot(data, aes(class, value)) +
    geom_jitter(aes(color = class), 
                height = 0.01, 
                width = 0.1, 
                alpha = 0.5, 
                shape = 20, size = 1) +
    geom_violin(alpha = 0.3) +
    ggsignif::geom_signif(comparisons = comparisons, family = font_family,
                          textsize = 3,
                          vjust = -.2,
                          step_increase = .18,
                          test=ttest2) +
    theme_bw() + 
    scale_y_continuous(expand = c(0.2, 0)) +
    ggtitle(title) +
    theme(axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()) +
    ylab("AUC Enrichment Score") +
    facet_wrap(~variable, 
               ncol = 3, 
               scales = "fixed")
  p
  
}


```

We then load the cell cycle signatures defined by McCarroll et al. and Glotzer et al,
calculate enrichment scores for each signature and each cell and plot results. 

```{r, results = 'hide', message = FALSE}

tab2sig <- function(file) {
  data <- read.delim(file, as.is=TRUE, stringsAsFactors = FALSE, header=TRUE)
  res <- apply(data, 2, function (x) {
    x <- gsub(" ", "", x)
    x <- x[which(x != "")]
  })
  res
}

cd <- readRDS("final_data/filtered_count_matrix.rds")
pd <- readRDS("final_data/cycling_metadata.rds")
sg <- pd$type_cdc20
sg <- sg[which(pd$barcode %in% colnames(cd))]
names(sg) <- colnames(cd)
sg <- factor(sg)

sg2 <- pd$type
sg2 <- sg2[which(pd$barcode %in% colnames(cd))]
names(sg2) <- colnames(cd)
sg2 <- factor(sg2)

auc_mccaroll <- function() {
  sigs <- tab2sig("final_data/signatures/cell_cycle_mccarroll.txt")
  cells_rankings <- AUCell_buildRankings(cd, nCores=12, plotStats=FALSE)
  cells_AUC <- AUCell_calcAUC(sigs, cells_rankings)
  dat <- as.data.frame(t(getAUC(cells_AUC)))
  dat$class <- sg
  dat <- melt(dat, id.vars = c("class"))
  dat$class <- factor(dat$class, labels = c("G1-CDC20-", 
                                        "G2/M-CDC20-", 
                                        "G2/M-CDC20+"))
  comparisons <- list( c("G2/M-CDC20+", "G2/M-CDC20-"),  
                   c("G1-CDC20-", "G2/M-CDC20-"), 
                   c("G1-CDC20-", "G2/M-CDC20+"))

  plot_auc(dat, comparisons = comparisons) +
    scale_color_discrete(labels=c(expression(paste("G1 CDC20"^low)),
                                  expression(paste("G1/M CDC20"^low)),
                                  expression(paste("G1/M CDC20"^high))))+
    theme(legend.text.align = 0) +
    labs(color="")
}

pdf(file="fig5a.pdf", width = 2000/300, height=1800/300, family = font_family)
p <- auc_mccaroll() +
      guides(color = guide_legend(override.aes = list(size = 4)))
print(p)
dev.off()

auc_glotzer <- function() {
  sigs <- tab2sig("final_data/signatures/CYTOKINESES_GLOTZER.txt")
  sigs <- as.data.frame(sigs)
  cells_rankings <- AUCell_buildRankings(cd, nCores=4, plotStats=FALSE)
  cells_AUC <- AUCell_calcAUC(sigs, cells_rankings)
  dat <- as.data.frame(t(getAUC(cells_AUC)))
  dat$class <- sg
  dat <- melt(dat, id.vars = c("class"))
  dat$class <- factor(dat$class, labels = c("G1-CDC20-", 
                                        "G2/M-CDC20-", 
                                        "G2/M-CDC20+"))
  comparisons <- list( c("G2/M-CDC20+", "G2/M-CDC20-"),  
                   c("G1-CDC20-", "G2/M-CDC20-"), 
                   c("G1-CDC20-", "G2/M-CDC20+"))
  plot_auc(dat, title=" ", comparisons = comparisons) + 
    scale_color_discrete(labels=c(expression(paste("G1 CDC20"^low)),
                                  expression(paste("G1/M CDC20"^low)),
                                  expression(paste("G1/M CDC20"^high))))+
    theme(legend.text.align = 0)
}

pdf(file="fig5d.pdf", width = 1000/300, height=900/300, family = font_family)
p <- auc_glotzer() +
    labs(color="") + 
    guides(color = guide_legend(override.aes = list(size = 4)))
print(p)
dev.off()

auc_chd4 <- function() {
  sigs <- tab2sig("chd4/chd_sh.txt")
  sigs <- as.data.frame(sigs$CHD4_sh_DOWN)
  cells_rankings <- AUCell_buildRankings(cd, nCores=4, plotStats=FALSE)
  cells_AUC <- AUCell_calcAUC(sigs, cells_rankings)
  dat <- as.data.frame(t(getAUC(cells_AUC)))
  dat$class <- sg
  dat <- melt(dat, id.vars = c("class"))
  dat$class <- factor(dat$class, labels = c("G1-CDC20-SPG20-", 
                                        "G2/M-CDC20-SPG20-", 
                                        "G2/M-CDC20+SPG20-",
                                        "G2/M-CDC20+SPG20+"))
  comparisons <- list( c("G2/M-CDC20+SPG20-", "G2/M-CDC20+SPG20+"),
                       c("G2/M-CDC20-SPG20-", "G2/M-CDC20+SPG20-")) 
  plot_auc(dat, comparisons = comparisons) +
    scale_color_discrete(labels=c(expression(paste("G1 CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^high)))) +
    theme(legend.text.align = 0) +
    labs(color="")
}
```

We can further stratify by SPG20 and repeat the above analysis.

```{r}
cd <- readRDS("final_data/filtered_count_matrix.rds")
pd <- readRDS("final_data/cycling_metadata.rds")
sg <- pd$type
sg <- sg[which(pd$barcode %in% colnames(cd))]
names(sg) <- colnames(cd)
sg <- factor(sg)

auc_mccaroll_spg <- function() {
  sigs <- tab2sig("final_data/signatures/cell_cycle_mccarroll.txt")
  cells_rankings <- AUCell_buildRankings(cd, nCores=12, plotStats=FALSE)
  cells_AUC <- AUCell_calcAUC(sigs, cells_rankings)
  dat <- as.data.frame(t(getAUC(cells_AUC)))
  dat$class <- sg
  dat <- melt(dat, id.vars = c("class"))
  dat$class <- factor(dat$class, labels = c("G1-CDC20-SPG20-", 
                                        "G2/M-CDC20-SPG20-", 
                                        "G2/M-CDC20+SPG20-",
                                        "G2/M-CDC20+SPG20+"))
  comparisons <- list( c("G2/M-CDC20+SPG20-", "G2/M-CDC20+SPG20+")) 
  plot_auc(dat, comparisons = comparisons) +
    scale_color_discrete(labels=c(expression(paste("G1 CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^high)))) +
    theme(legend.text.align = 0) +
    labs(color="")
}

pdf(file="fig6g.pdf", width = 2200/300, height=1800/300, family = font_family)
p <- auc_mccaroll_spg() + 
      guides(color = guide_legend(override.aes = list(size = 4)))
print(p)
dev.off()

auc_glotzer_spg <- function() {
  sigs <- tab2sig("final_data/signatures/CYTOKINESES_GLOTZER.txt")
  sigs <- as.data.frame(sigs)
  cells_rankings <- AUCell_buildRankings(cd, nCores=4, plotStats=FALSE)
  cells_AUC <- AUCell_calcAUC(sigs, cells_rankings)
  dat <- as.data.frame(t(getAUC(cells_AUC)))
  dat$class <- sg
  dat <- melt(dat, id.vars = c("class"))
  dat$class <- factor(dat$class, labels = c("G1-CDC20-SPG20-", 
                                        "G2/M-CDC20-SPG20-", 
                                        "G2/M-CDC20+SPG20-",
                                        "G2/M-CDC20+SPG20+"))
  comparisons <- list( c("G2/M-CDC20+SPG20-", "G2/M-CDC20+SPG20+")) 
  plot_auc(dat, comparisons = comparisons) +
    scale_color_discrete(labels=c(expression(paste("G1 CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^low*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^low)),
                                  expression(paste("G2/M CDC20"^high*"SPG20"^high)))) +
    theme(legend.text.align = 0)
}

set.seed(1000)
p <- auc_glotzer_spg() +
    labs(color="") + 
    guides(color = guide_legend(override.aes = list(size = 4)))
pdf(file="fig6h.pdf", width = 1300/300, height=900/300, family = font_family)
#png(file="fig6h.png", res = 300, width = 1300, height = 900)
print(p)
dev.off()

```

